---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(lme4)

simdat <- read.csv(file.choose())
```

# Challenge questions

Using the binomial (log linear) model in `lme4`, construct and interpret a model that examines accuracy as a function of age, frequency, grammaticality, and the interaction of frequency and grammaticality. Include random effects for intercept and figure out if any random effects for slope will converge.

```{r challenge-one, warning=FALSE}
# maximal model
# measure ~ factor1 + factor2 + factor1:factor2 + (1 + random slopes | random intercept)

challenge.mdl.max <- glmer(accuracy ~ 1 +
                            age + 
                            freq + 
                            gram + 
                            freq:gram +
                            (1 | subj) +
                            (1 | item),
                          data = simdat %>% filter(region==1),
                          family = "binomial")
summary(challenge.mdl.max)
```

```{r}
# add depleted models to test the contribution of each fixed effect
challenge.mdl.int <- glmer(accuracy ~ 1 +
                            age + 
                            freq + 
                            gram + 
                            #freq:gram +
                            (1 | subj) +
                            (1 | item),
                          data = simdat %>% filter(region==1),
                          family = "binomial")

anova(challenge.mdl.max, challenge.mdl.int)
```

```{r}
challenge.mdl.frq <- glmer(accuracy ~ 1 +
                            age + 
                            #freq + 
                            gram + 
                            #freq:gram +
                            (1 | subj) +
                            (1 | item),
                          data = simdat %>% filter(region==1),
                          family = "binomial")
anova(challenge.mdl.frq, challenge.mdl.int)
```

```{r}
challenge.mdl.grm <- glmer(accuracy ~ 1 +
                            age + 
                            freq + 
                            #gram + 
                            #freq:gram +
                            (1 | subj) +
                            (1 | item),
                          data = simdat %>% filter(region==1),
                          family = "binomial")
anova(challenge.mdl.grm, challenge.mdl.int)
```

```{r}
challenge.mdl.age <- glmer(accuracy ~ 1 +
                            #age + 
                            freq + 
                            gram + 
                            freq:gram +
                            (1 | subj) +
                            (1 | item),
                          data = simdat %>% filter(region==1),
                          control = glmerControl(optimizer = "bobyqa"),
                          family = "binomial")
anova(challenge.mdl.age, challenge.mdl.max) %>% broom::tidy()
```


To help you interpret the interaction term, here is a plot of the interaction (without age or random effects accounted for).

```{r, message=FALSE}
simdat %>% 
  filter(region==1) %>% 
  group_by(age,freq,gram) %>% 
  summarise(n=n(),
            acc=sum(accuracy),
            correct= acc/n) %>% 
  ggplot(aes(x = freq, colour=gram, y = correct)) +
  geom_point() +
  geom_path(aes(group=gram)) +
  theme_bw() +
  facet_wrap(~age) +
  NULL

simdat %>% 
  filter(region==1) %>% 
  group_by(freq,gram) %>% 
  summarise(n=n(),
            acc=sum(accuracy),
            correct= acc/n) %>% 
  ggplot(aes(x = freq, colour=gram, y = correct)) +
  geom_point() +
  geom_path(aes(group=gram)) +
  theme_bw() +
  NULL
```

