---
title: "Before the workshop..."
author: "Dr Lauren Ackerman"
output:
  html_document:
    toc: false
    number_sections: false
    theme: sandstone
    highlight: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set up

Please make sure you have downloaded and installed [R](https://cran.r-project.org/), [RStudio](https://www.rstudio.com/products/rstudio/download/), and these packages below on your computer **before the workshop**. You may contact me if you run into problems. 

```{r eval=FALSE, eval=FALSE}
install.packages("tidyverse","ordinal","lme4","broom")
```

If you are unfamiliar with R and RStudio, follow these instructions:

1. Download and install R and RStudio on the computer you will be using for this workshop.
2. Locate on your computer the file called `20210608-datawrangling.Rmd` that was sent to you in advance of the workshop.
3. Open that file in RStudio. You will now be looking at a white and grey document.
4. At the top right of the second grey area, there should be a green triangle pointing rightwards. Click it.
5. Allow the interface to work its magic! (Red text isn't necessarily bad -- it can simply be information.)

This workshop will assume some familiarity with both R and RStudio. If you are entirely new to R, RStudio, programming and/or statistics, please also install the following package by clicking the green triangle in the upper right corner of the grey area (when viewing this file in RStudio):

```{r, eval=FALSE}
install.packages("swirl")
```

The `swirl` package will give you some basic skills that you will need for the workshop, if you do not already have R experience.

Run the following chunk as before (green triangle) to get started with `swirl`.
```{r, eval=FALSE}
swirl::swirl()
```

Once you have clicked the green triangle, you will see it becomes a red square. You may also see a 'waiting' symbol moving below the chunk. If this is all you see (that is, you do not see red text saying 'Welcome to swirl! ...'), find and click the `Console` tab in the RStudio window. You will interact with `swirl` through the console.

To exit `swirl`, you can press the Escape key.

# See you soon!