---
title: "**Simulated data**"
author: "Dr Lauren M Ackerman"
date: "28 FEB 2019"
output:
  html_notebook:
    toc: false
    theme: yeti
    highlight: default
---

```{r}
library(lme4)
library(tidyverse)
```

# Fake Experiment {.tabset}

Kinds of data:

|Self-paced reading | acceptability | accuracy|
|-------------------|---------------|---------|
|continuous (rt) | ordinal (Likert) | binomial (yes/no)|

Stimulus design:

|Condition | Sample sentence | Verb frequency | Grammaticality |
|----------|-----------------|----------------|----------------|
|1|the old man the boat|high|  grammatical|
|2|the old put the boat|high|ungrammatical|
|3|the old run the boat|low |  grammatical|
|4|the old owe the boat|low |ungrammatical|

## Set up objects

```{r}
subj <- rep(c(1:40),each=20)
set.seed(2019); ages <- floor((rpois(40,1)+5+rnorm(40,mean=5,sd=3))*4)
  age <- rep(ages,each=20)
item <- rep(c(1:20),40)
set.seed(2019); itemDifficulty <- rep(rnorm(20,mean=0,sd=1),40)
freq <- rep(c("high","low"),40,each=10)
gram <- rep(c("yes","no"),80,each=5)

df.rt <- tibble(subj,age,item,freq,gram,itemDifficulty)


df.rt
```

## Create dataset

```{r}
set.seed(2019); df.rt %>%
  mutate(reg1.the  = rnorm(800, mean = 320, sd=100)+age+itemDifficulty*2,
         reg2.old  = rnorm(800, mean = 360, sd=80)+age +itemDifficulty*2,
         reg3.VERB = rnorm(800, mean = 370, sd=80)+age +itemDifficulty*2,
         reg4.the  = rnorm(800, mean = 320, sd=100)+age+itemDifficulty*2,
         reg5.boat = rnorm(800, mean = 370, sd=150)+age+itemDifficulty*2) %>%
  mutate(reg3.VERB = case_when(freq=="high" ~ reg3.VERB-rnorm(200,35,5),
                               TRUE ~ reg3.VERB)) %>%
  mutate(reg4.the  = case_when(gram=="no" ~ reg4.the+rnorm(200,20,10),
                              TRUE ~ reg4.the)) %>%
  mutate(reg5.boat = case_when(gram=="no" ~ reg5.boat+rnorm(200,45,5),
                              TRUE ~ reg5.boat)) %>%
  mutate(rating = case_when(freq=="high" & 
                              gram=="yes" ~ sample(x   =c(1,   2,  3,   4,   5),
                                                   prob=c(0.01,0.1,0.14,0.25,0.5),
                                                   size=800,
                                                   replace=TRUE),
                            freq=="low"  & 
                              gram=="yes" ~ sample(x   =c(1,   2,  3,   4,   5),
                                                   prob=c(0.01,0.11,0.13,0.26,0.49),
                                                   size=800,
                                                   replace=TRUE),
                            freq=="high" & 
                              gram=="no"  ~ sample(x   =c(1,   2,  3,   4,   5),
                                                   prob=c(0.3,0.26,0.27,0.12,0.04),
                                                   size=800,
                                                   replace=TRUE),
                            freq=="low"  & 
                              gram=="no"  ~ sample(x   =c(1,   2,  3,   4,   5),
                                                   prob=c(0.5,0.25,0.14,0.1,0.01),
                                                   size=800,
                                                   replace=TRUE))) %>%
  mutate(accuracy = case_when(gram=="no" & freq=="high" & age < 35 ~            
                                sample(c(0,1),prob = c(0.1,0.9), size=800, replace=TRUE),
                              gram=="no" & freq=="high" & age >=35 & age <=55 ~ 
                                sample(c(0,1),prob = c(0.15,0.85), size=800, replace=TRUE),
                              gram=="no" & freq=="high" & age > 55 ~            
                                sample(c(0,1),prob = c(0.3,0.7), size=800, replace=TRUE),

                              gram=="no" & freq=="low"  & age < 35 ~            
                                sample(c(0,1),prob = c(0.05,0.95), size=800, replace=TRUE),
                              gram=="no" & freq=="low"  & age >=35 & age <=55 ~ 
                                sample(c(0,1),prob = c(0.1,0.9), size=800, replace=TRUE),
                              gram=="no" & freq=="low"  & age > 55 ~            
                                sample(c(0,1),prob = c(0.25,0.75), size=800, replace=TRUE),

                              gram=="yes" & freq=="high" & age < 35 ~            
                                sample(c(0,1),prob = c(0.02,0.98), size=800, replace=TRUE),
                              gram=="yes" & freq=="high" & age >=35 & age <=55 ~ 
                                sample(c(0,1),prob = c(0.05,0.95), size=800, replace=TRUE),
                              gram=="yes" & freq=="high" & age > 55 ~            
                                sample(c(0,1),prob = c(0.2,0.8), size=800, replace=TRUE),

                              gram=="yes" & freq=="low" & age < 35 ~            
                                sample(c(0,1),prob = c(0.05,0.95), size=800, replace=TRUE),
                              gram=="yes" & freq=="low" & age >=35 & age <=55 ~ 
                                sample(c(0,1),prob = c(0.1,0.9), size=800, replace=TRUE),
                              gram=="yes" & freq=="low" & age > 55 ~            
                                sample(c(0,1),prob = c(0.25,0.75), size=800, replace=TRUE))) %>%
  gather(region,rt,7:11) %>%
  separate(col=region,into=c("region","word")) %>%
  mutate(region = as.numeric(substr(region,4,4))) %>%
  select(-itemDifficulty) -> data
```

## Write to file

```{r}
write.csv(data, file = "../data/simulated-data.csv", row.names = FALSE)
```

