---
title: "Life in the UK"
date: 2021-04-11
output: 
  html_document:
    df_print: paged
    theme: sandstone
    highlight: default
    toc: false
    includes: 
      in_header: google_analytics.html
---

```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

```{r setup, include=FALSE}
# install.packages("vistime")
library(vistime); library(tidyverse)
```

## Introduction

In this notebook, I have made a series of timeline plots from the same data, presented below. This spreadsheet was made from the Life in the UK handbook (and additional information to contextualise the timing of relevant events):

```{r data, echo=FALSE}
# century = 36525 days
timedat <- read.csv("data/timeline.csv")

timedat %>% as_tibble() %>% 
  mutate(start = as.Date(as.numeric(start), origin="0000-01-01"),
         end = as.Date(as.numeric(end), origin="0000-01-01"),
         altStart = as.Date(as.numeric(altStart), origin="0000-01-01"),
         altEnd = as.Date(as.numeric(altEnd), origin="0000-01-01")) -> timeline
data.table::data.table(timeline %>% 
                         transmute(Era = group,
                                   Event = event,
                                   Start = startForHumans,
                                   End = endForHumans,
                                   `Start date` = start,
                                   `End date` = end) %>% 
                         arrange(`Start date`))
```

Not all of the dates are completely accurate. Some are taken from general dates listed in the official handbook, others are from wikipedia or other online sources. The purpose of these dates is primarily to rearrange the information presented in the official handbook in chronological order, to better visualise when different events occurred.

</hr>

Each of the tabs below provides a slightly different method for generating the timelines.

# `vistime` package {.tabset}

## `gg_vistime()` 

This method produces a static visualisation using ggplot functions.

> Instructions: click on plot to zoom in to full size. click again to exit.

### whole timeline (gg)

```{r ggvis, echo=FALSE, fig.dim=c(20,10)}
gg_vistime(timeline %>% 
             mutate(start=altStart,end=altEnd) %>% 
             #select(-color) %>% 
             arrange(start), 
           optimize_y = FALSE,
           linewidth = 5)
#gg_vistime(timeline %>% select(-color), optimize_y = FALSE)
```

### common era only (gg)

```{r cegg, echo=FALSE, fig.dim=c(15,7)}
gg_vistime(timeline %>% 
             mutate(start=altStart,end=altEnd) %>% 
             #select(-color) %>% 
             filter(start >= "0000-01-01") %>% 
             arrange(start), 
           optimize_y = FALSE,
           background_lines = 25) 
```

## `vistime()` 

This method is the default plotly visualisation.

> Instructions: hover over image to see plotly navigation options.

### whole timeline

```{r vistimeplotly, echo=FALSE, warning=FALSE, message=FALSE, fig.dim=c(7,5)}
vistime(timeline %>% 
          mutate(start=altStart, end=altEnd) %>% 
          #select(-c(color,tooltip)) %>% 
          arrange(start), 
        optimize_y = FALSE)
```

- I dont know why the Bronze Age and other events don't show up

### common era only

```{r cebase, echo=FALSE,message=FALSE}
vistime(timeline %>% 
          mutate(start=altStart, end=altEnd) %>% 
          #select(-c(color,tooltip)) %>% 
          filter(start >= "0000-01-01") %>% 
          arrange(start), 
        optimize_y = FALSE)
```


## `hc_vistime()` 

This method produces an interactive visualisation with fewer view options than the default, but a cleaner looking plot.

> Instructions: hover over image to see interactive elements. click and drag to zoom in to a selected area.

### whole timeline (hc)

```{r hcvis, echo=FALSE, warning=FALSE, message=FALSE}
hc_vistime(timeline %>% 
             mutate(start=altStart,end=altEnd) %>% 
             select(-color) %>% 
             arrange(start), 
           optimize_y = FALSE)
#hc_vistime(timeline %>% select(-color), optimize_y = FALSE)
```

### common era only (hc)

```{r cehc, echo=FALSE, message=FALSE}
hc_vistime(timeline %>% 
             select(-color) %>% 
             filter(start >= "0000-01-01") %>% 
             arrange(start), 
           optimize_y = FALSE)
```

