---
title: "Exploration and Troubleshooting"
output: 
  html_document:
    theme: cerulean
    highlight: default
    toc: true
    toc_depth: 2
    number_sections: true
    toc_float: 
      smooth_scroll: true
      collapsed: false
---
[⇠ Turorial 2](tutorial-2.html)

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("header.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:20px;')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

|  | Learning Outcomes |
|--|----------|
| 1|<span style="background-color: #FFFF00">You will know how to plan out code in a step-wise manner</span>|
| 2|<span style="background-color: #FFFF00">You will be able to navigate the CRAN documentation pages</span>|
| 3|<span style="background-color: #FFFF00">You will know how to search online for solutions</span>|
| 4|<span style="background-color: #FFFF00">You will realise that being good at R is just being good at Google</span>|
|||


# Planning code

In order to write effective code in any language, you must start to think like a programmer. This is a lot easier than it might sound. Programmers approach problems in a step-wise manner, which allows them to figure out and plan each step of the solution. When they come up against a step that they can't implement right away, it might indicate that the step is actually too complex as written out and needs to be broken into smaller chunks. It's also possible that it's the right sized chunk, but they just don't know the function to accomplish the step. In that case, there is probably a specific search term they can use to figure out how other people may have dealt with the same issue. In either case, this type of workflow will help you feel more comfortable with using R or any computer programming language you might be interested in.

## Strategise

## Test

## Run

# CRAN documentation
Reading CRAN documentation

## Formulas

## Arguments

## Examples

# Online resources
Intro to GitHub and StackOverflow

## StackOverflow

### Searching for answers

### Asking questions

### Pitfalls

## GitHub

### What is GitHub?

### Basics
What do you actually need to know to make use of GitHub?

### Getting help
