---
title: "Exploration and Troubleshooting"
output: 
  html_document:
    theme: cerulean
    highlight: default
    toc: true
    toc_depth: 2
    number_sections: true
    toc_float: 
      smooth_scroll: true
      collapsed: false
---
[⇠ Turorial 2](tutorial-2.html)

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("header.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:20px;')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

|  | Learning Outcomes |
|--|----------|
| 1|<span style="background-color: #FFFF00">You will know how to plan out code in a step-wise manner</span>|
| 2|<span style="background-color: #FFFF00">You will be able to navigate the CRAN documentation pages</span>|
| 3|<span style="background-color: #FFFF00">You will know how to search online for solutions</span>|
| 4|<span style="background-color: #FFFF00">You will realise that being good at R is just being good at Google</span>|
|||


# Planning code
Thinking like a programmer

## Strategise

## Test

## Run

# CRAN documentation
Reading CRAN documentation

## Formulas

## Arguments

## Examples

# Online resources
Intro to GitHub and StackOverflow

## StackOverflow

### Searching for answers

### Asking questions

### Pitfalls

## GitHub

### What is GitHub?

### Basics
What do you actually need to know to make use of GitHub?

### Getting help
