---
title: "Intro to Tidyverse"
output: 
  html_document:
    theme: cerulean
    highlight: default
    toc: true
    number_sections: true
    toc_float: 
      smooth_scroll: true
      collapsed: false
---
[⇠ Turorial 1](tutorial-1.html) || [Tutorial 3 ⇢](tutorial-3.html)

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("header.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:20px;')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

|  | Learning Outcomes |
|--|----------|
| 1|You will be able to load tidyverse and use basic functions|
| 2|You will be able to relabel and reorganise a dataset|
| 3|You will be able to summarise a dataset|
| 4|You will be able to generate a simple plot|
|||


# What is Tidyverse?

```{r}
library(tidyverse)
```


# Basic data manipulation

## Piping

## Mutate

## Filter

## Transmute

## Gather / Melt

## Spread

# Summarising a data table

## Group by

## Summarise

### Count

### Mean

### Standard deviation

### Custom value

# Ingredients of a ggplot

## Base plot

### Aesthetics

#### X-axis

#### Y-axis

#### Colour

#### Fill

#### Size

#### Alpha

## Geom

### Point

### Bar

### Boxplot

### Hist

### Smooth
