---
title: "File management and datasets"
output: 
  html_document:
    theme: cerulean
    highlight: default
    toc: true
    number_sections: true
    toc_float: 
      smooth_scroll: true
      collapsed: false
---
[⇠ RfficeHours Home](info.html) || [Tutorial 2 ⇢](tutorial-2.html)

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("header.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; padding:20px;')
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# File structure
Navigating file structures in RStudio

## R vs RStudio
What is the difference between R and RStudio?

## Scripts
What is the point of a script?

## Best practices
Best practices for file organisation

# Reading and writing
Reading and writing files

## Scripts
Opening and creating scripts

## Datasets
Opening datasets

### Tab-delimited files (.txt)

### Comma separated values (.csv)

### Why not Excel? (.xls/.xlsx)

# Literate programming
Rmarkdown and literate programming

## Annotation
Why do we annotate code?

## Prose
What is literate programming?

## R Notebooks

### Organising code

### Sharing code

### Publishing code